---
title: "DESeq2"
output: html_document
---

```{r setup and packages, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DESeq2)
library(tidyverse)
library(readxl)
library(BiocManager)
library("BiocParallel")
library("pheatmap")
library(dplyr)
library(EnsDb.Hsapiens.v79)
register(SnowParam(4))
library(VennDiagram)
library(gplots)
library(RColorBrewer)
library("ggplot2")



```
 Import data

```{r import_data}
matrix <- read.table("../Desktop/genes.tvs/genes.tsv", header=1, check.names = FALSE)
rownames(matrix) <- matrix$id
matrix$id <- NULL


metadata <- read_excel("C:/Users/argyr/Desktop/metadata.xlsx")
metadata$timepoint<-as.factor(metadata$timepoint)


dds<-DESeqDataSetFromMatrix(countData = matrix , colData = metadata, design = ~timepoint+donor)
dds

```
gene names

```{r names}
#   BiocManager::install("EnsDb.Hsapiens.v79")

ensembl.genes <- rownames(res)
geneIDs1 <- ensembldb::select(EnsDb.Hsapiens.v79, keys= ensembl.genes, keytype = "GENEID", columns = c("SYMBOL","GENEID"))


counts <- as.data.frame(counts(dds, normalized=T))
counts$GENEID <- rownames(counts)
combine_counts <- dplyr::inner_join(counts, geneIDs1, by="GENEID")
rownames(combine_counts) <- make.unique(combine_counts$SYMBOL)
combine_counts$GENEID <- NULL
combine_counts$SYMBOL <- NULL
```

Dataset with gene names
```{r dataset with names}
count$id <- rownames(count) 
count$id <- NULL

dds1<-DESeqDataSetFromMatrix(countData = combine_counts , colData = metadata, design = ~timepoint) 
dds1
```

LRT significant - gene names
``` {r }

ddsLRT <- DESeq(dds, test="LRT",reduced= ~ timepoint)
ddsLRT
DEresLRT <- results(ddsLRT)
DEresLRT
summary(DEresLRT, alpha=0.05)
resLRT <- as.data.frame(DEresLRT) + dplyr::filter(padj < 0.05)
resLRT

# or

ddsLRT <- DESeq(dds1, test="LRT",reduced= ~ 1)
ddsLRT
DEresLRT <- results(ddsLRT)
DEresLRT
summary(DEresLRT, alpha=0.05)
resLRT <- as.data.frame(DEresLRT)   %>% rownames_to_column('id')  %>%   dplyr::filter(padj > 0.05)
view(resLRT)
```
 Differential expression analysis
```{r Differential expression analysis}
dds <- DESeq(dds)
dds1 <- DESeq(dds1)
results(dds)

res02<- as.data.frame(results(dds1, contrast = c("timepoint", "2", "0"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res02) # +1 TMEM189-UBE2V1 ENSG00000124208

res04<- as.data.frame(results(dds1, contrast = c("timepoint", "4", "0"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res04)

res06 <- as.data.frame(results(dds1, contrast = c("timepoint", "6", "0"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res06)

res08 <- as.data.frame(results(dds1, contrast = c("timepoint", "8", "0"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res08)

res24<- as.data.frame(results(dds1, contrast = c("timepoint", "2", "4"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05)
view(res24)

res48<- as.data.frame(results(dds1, contrast = c("timepoint", "4", "8"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05)
view(res48)

res46<- as.data.frame(results(dds1, contrast = c("timepoint", "4", "6"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res46)

res68<- as.data.frame(results(dds1, contrast = c("timepoint", "6", "8"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res68)

res46<- as.data.frame(results(dds1, contrast = c("timepoint", "4", "6"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res46)

res48<- as.data.frame(results(dds1, contrast = c("timepoint", "4", "8"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res48)


res26<- as.data.frame(results(dds1, contrast = c("timepoint", "2", "6"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res26)

res28<- as.data.frame(results(dds1, contrast = c("timepoint", "2", "8"))) %>% rownames_to_column('id')  %>% dplyr::filter(padj < 0.05) 
view(res28)
```

LRT significant - gene names
``` {r }
ddsLRT <- DESeq(dds1, test="LRT",reduced= ~ 1)
ddsLRT
DEresLRT <- results(ddsLRT)
DEresLRT
summary(DEresLRT, alpha=0.05)
resLRT <- as.data.frame(DEresLRT)   %>% rownames_to_column('id')  %>%   dplyr::filter(padj > 0.05)
view(resLRT)
```
Differential expression analysis
```{r Differential expression analysis}

counts <- as.data.frame(counts(dds, normalized=TRUE))
counts$GENEID <- rownames(counts)
combine_counts <- dplyr::inner_join(counts, geneIDs1, by="GENEID")
rownames(combine_counts) <- make.unique(combine_counts$SYMBOL)
combine_counts$GENEID <- NULL
combine_counts$SYMBOL <- NULL

dds <- DESeq(dds)
res <- results(dds, contrast = c("timepoint", "2", "0"))
resultsNames(dds)
res
```

tfs
```{r tfs}
tf<- read.table("../Desktop/export.txt", sep="\t" , header=1, check.names = FALSE)
tf$Human_entrez <- NULL
tf$Rat_entrez <- NULL
tf$Mouse_entrez<- NULL
tf$Gene_Name <- NULL
tf$Synonym <- NULL
tf$DbTF <- NULL
tf$PMID <- NULL
tf$Uniprot_Rat <- NULL
tf$Uniprot_mouse <- NULL
tf$Uniprot_Human <- NULL
tf

#Dataset 0-8 significant

null8 <- read_excel("C:/Users/argyr/Desktop/08_sig.xlsx")
null8


common <- intersect(null8$Genes, tf$Gene_symbol)
common

tf08<-  c("NR1H3"  , "ARNTL2"  ,"PRDM1" ,  "YBX3" ,   "RARB"    ,"TCF7"  ,  "TBX15"  , "RELB" ,  
  "HLF"    , "BCL6"  ,  "NFE2L2",  "BHLHE41" ,"HIF3A" ,  "NR1D1" ,  "RNF112"  ,"HMGA1"  ,
 "LEF1"   , "OSR1" ,   "THRB" ,   "ZNF281" , "ZNF219" , "E2F7" ,   "MN1"  ,   "SP7" ,   
 "STAT2" ,  "STAT5B" , "FOSL1" ,  "MYBL1" ,  "LITAF"   )
tf08


```
##expression in time/up down regulation
``` {r whatever}
results(dds1)
reg<- as.data.frame(results(dds1)) %>% rownames_to_column('id') 
reg 
upreg<-   arrange(reg, desc(log2FoldChange)) %>%
 dplyr::filter( padj < 0.05) %>% dplyr::filter(log2FoldChange > 0)
upreg
downreg<-arrange(reg, log2FoldChange) %>%
 dplyr::filter( padj < 0.05) %>% dplyr::filter(log2FoldChange < 0)
downreg


common1 <- intersect(upreg$id, downreg$id)  
common1 
```

Heatmap with all donors, all timepoints.
Genes that significantly change from 0 to 8.

```{r heatmap}

select<- order(rowMeans(combine_counts), decreasing = TRUE)
df<-as.data.frame(colData(dds1)[,c("donor","timepoint")])
ntd <- normTransform(dds1)

mat <- assay(ntd) %>% as.data.frame() %>% 
  tibble::rownames_to_column("GENES") %>% 
  dplyr::filter(GENES %in% null8$Genes) %>% 
  tibble::column_to_rownames("GENES")
mat <- as.matrix(mat)
mat <- mat - rowMeans(mat)

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)
```


Heatmap that includes the genes that change significantly from 0 to 8.
Heatmap includes all donors, only timepoints 0 and 8.


```{r}

timepoint08order<- subset(combine_counts, select=c(5, 11, 1, 17 , 10, 15 , 16, 2)) 

metadata <- read_excel("C:/Users/argyr/Desktop/metadata1.xlsx")
metadata$timepoint<-as.factor(metadata$timepoint)
dds2<-DESeqDataSetFromMatrix(countData = timepoint08order , colData = metadata, design = ~timepoint) 
dds2
dds2<-DESeq(dds2)

select<- order(rowMeans(timepoint08order), decreasing = TRUE)
df<-as.data.frame(colData(dds2)[,c("donor","timepoint")])
ntd <- normTransform(dds2)

mat <- assay(ntd) %>% as.data.frame() %>% 
  tibble::rownames_to_column("GENES") %>% 
  dplyr::filter(GENES %in% null8$Genes) %>% 
  tibble::column_to_rownames("GENES")
mat <- as.matrix(mat)
mat <- mat - rowMeans(mat)

mat_colors <- list(group = brewer.pal(3, "Set1"))
pheatmap(mat, cluster_rows=TRUE,color = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdBu")))(20), show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)

```
Heatmap with all donors, all timepoints.
Genes are the TFs

```{r}


select<- order(rowMeans(combine_counts), decreasing = TRUE)
df<-as.data.frame(colData(dds1)[,c("donor","timepoint")])
ntd <- normTransform(dds1)



mat <- assay(ntd) %>% as.data.frame() %>% 
  tibble::rownames_to_column("GENES") %>% 
  dplyr::filter(GENES %in% tf08) %>% 
  tibble::column_to_rownames("GENES")
mat <- as.matrix(mat)
mat <- mat - rowMeans(mat)

mat_colors <- list(group = brewer.pal(3, "Set1"))
pheatmap(mat, cluster_rows=TRUE,color = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdBu")))(20), show_rownames=TRUE, cluster_cols=TRUE, annotation_col=df)
```
Heatmap with all donors, timepoints 0 to 8.
Genes are the TFs

```{r}


select<- order(rowMeans(timepoint08order), decreasing = TRUE)
df<-as.data.frame(colData(dds2)[,c("donor","timepoint")])
ntd <- normTransform(dds2)



mat <- assay(ntd) %>% as.data.frame() %>% 
  tibble::rownames_to_column("GENES") %>% 
  dplyr::filter(GENES %in% tf08) %>% 
  tibble::column_to_rownames("GENES")
mat <- as.matrix(mat)
mat <- mat - rowMeans(mat)

mat_colors <- list(group = brewer.pal(3, "Set1"))
pheatmap(mat, cluster_rows=TRUE,color = colorRampPalette(rev(brewer.pal(n = 7, name =
  "RdBu")))(20), show_rownames=TRUE, cluster_cols=FALSE, annotation_col=df)
```


Plots for the expression of the TFs
```{r}

for(gene in tf08){
  f<-plotCounts(dds1, gene, intgroup="timepoint", returnData = TRUE)
  print(ggplot(f, aes( x=timepoint, y=count)) + geom_boxplot() + geom_text(aes(label = row.names(f)))+   ggtitle(gene)+ geom_point(position=position_jitter(w=0.1,h=0)) + scale_y_log10(breaks=waiver()))
}



```
Plots for the expression of the genes that we picked based on literature (Tas' plots)
```{r}

impgenes <-c("DMP1", "SOST", "PHEX", "MEPE", "RUNX2", "OMD", "ALPL", "BGLAP", "GDF5", "GJA1", "PDPN", "FGF23", "PTH1R", "TNFRSF11B","TNFSF11")

for(gene in impgenes){
  f<-plotCounts(dds1, gene, intgroup="timepoint", returnData = TRUE)
  print(ggplot(f, aes( x=timepoint, y=count)) + geom_boxplot() + geom_text(aes(label = row.names(f)))+   ggtitle(gene)+ geom_point(position=position_jitter(w=0.1,h=0)) + scale_y_log10(breaks=waiver()))
}



```

Venn diagram
```{r}

tf <- tf08
up <- upreg$id
down <- downreg$id
sigall <- null8$Genes


myCol <- brewer.pal(3, "Pastel2")

venn <- venn.diagram(
  x = list(tf, up, down),
  category.names = c("TFs" , "Up-regulated", "Down regulated") ,
  filename = 'venn.png' ,

  output = TRUE , 
  imagetype = "png",
          col=c("#440154ff", '#21908dff', '#fde725ff'),
          fill = myCol,
          cat.col = c("#440154ff", '#21908dff', '#fde725ff'),
  
                  # Numbers
        cex = 1,
        fontface = "bold",
        fontfamily = "sans",
        
        # Set names
        cat.cex = 1,
        cat.fontface = "bold",
        cat.default.pos = "outer",
        cat.fontfamily = "sans"
)


```